An agent-based model of military mechanization

### 2018-07-14 paper2/paper2.tex
justification

	Military mechanization is the degree to which a state's army is comprised of
	ground combat vehicles as opposed to infantry. Mechanization is both a 
	theoretically important topic in international relations scholarship and a practical 
	concern for policymakers. \citet{sechser2010army} influentially argue that
	inter-state dynamics (e.g., conflict) better explain mechanization choices than do
	intra-state factors (e.g., culture). This paper demonstrates that
	the multivariate statistical approach employed in \citet{sechser2010army} is
	inadequate and that agent-based modeling is needed to capture
	international security competition as a complex system. Candidate
	agent-based models are presented.

@@

Why ABM:
 - inadequacy of SS - VIF, linear model
 - uncertainty of causation
 - endogeneity
 - GDP / capita explains this...but also everything else
 - have to go with simulation
 - bigger picture, world politics as complex system
 -

Things they don't capture: selection into war and deterrence, gdp / capita to
political structure, institutional culture and choice, network effects and
emulation

Two central claims of this paper are that the multivariate statistical approach
of \citet{sechser2010army} is inadequate and agent-based modeling is needed.
They are addressed in turn.

\citet{sechser2010army} gather observational panel data that consists of a measure 
of military mechanization $y$ indexed by state $i$ and observation time $k$, as well as
time-indexed covariates $x$. The association of covariates with mechanization is
estimated via ordinary least squares (OLS) linear regression:

\begin{equation}
	y_{ik} = \beta_0 + \beta_j x_{ijk} + \epsilon \hspace{1em}
\end{equation}

Multicollinearity - everything correlated with each other.

The fitted model attempts to characterize the separate
effect of each factor upon mechanization. However, evidence suggests that
the covariates in fact are strongly cross-correlated. The problem of
multi-collinearity can be expressed both

@@

This paper presents an agent-based model of military mechanization. It
reviews the academic literature on military mechanization. It discusses why
orthodox modeling strategies fail to capture international security competition
as a complex system and why agent-based modeling is needed. In support of these
claims, it ``agentizes'' the leading model of military mechanization and shows that the
results of doing cast doubt on it. Finally, it presents a
novel agent-based model that better represents military mechanization.

In summary, the orthodox view of mechanization is that it is a
reflexive process: states choose their mechanization posture largely based on
that neighbors, enemies, and recent military experiences (e.g. defeat in
counter-insurgency). This paper argues
that this view---based on traditional longitudinal data analysis---is ripe for
agentization. It implicitly argues for autonomous agents (states) that have
cognition regarding mechanization that takes into account other agents (e.g.
enemies), spatiality (e.g. neighbors), and learning from experience. This paper
builds such a model and validates it against the empirical record, using the
National Mechanization Index of \citet{sechser2010army}. It finds that the 
orthodox view of mechanization does not stand up to agentization. A positive feedback loop 
causes states to
mechanize much faster and to a greater degree than occurs in the real-world.
This implies the existence of some countervailing factor (possibly a negative
feedback loop) that mitigates the extent of mechanization. Candidates for what
this could be mostly center on ``internal'' factors---e.g., culture, economy,
and so on---as opposed to the ``external'', security environment-centric
orthodox view.

@@

There are many candidates for the causes of mechanization. Inter-state
competition and war is one. States may choose choose mechanization policies 
that they believe will allow them to prevail against potential adversaries, or to send a
signal of deterrence, or to learn from the perceived mistakes of the last war
\citep{mearsheimer1983conventional,huth1988extended,murray2011military}. 
Economic endowments may play a role: e.g., capital-rich states may mechanize more
than labor-rich states \citep{gartzke2001democracy}. Domestic political
institutions may matter: e.g., autocracies may be more willing than democracies to adopt
low-mechanization ``human wave'' tactics
\citep{quinlivan1999coup,reiter2002democracies,brooks2008shaping,talmadge2015dictator}. 
Ideology \citep{van1984cult} and national culture \citep[e.g.][]{pollack2004arabs} are
hard-to-measure but perhaps influential. Lastly,
organizational culture can play a role \citep{builder1989masks}. Militaries may pursue a
certain mechanization posture because it aligns with their institutional
objectives, which may or may not be coincident with rational strategy. 
Consider Ireland: between the 1920s and 1940s, it seriously considered expending a 
large portion of its defense budget on building a relatively small number of tanks. 
They would have been wholly insufficient to repel a British invasion (their ostensible purpose), 
and Ireland would have been better-served to pursue a low-mechanization guerrilla defense strategy, 
but tanks were seen as a prestigious signifier of a ``real'' professional
military and hence desirable
\citep{farrell1998professionalization,farrell2001transnational}.

\citet{sechser2010army} is the landmark study on military mechanization. They
assemble a longitudinal dataset at the country-year level of the mechanization
of all states between 1979 and 2001. They conduct regression analysis to identify
the effect of covariates on mechanization. They conclude that out of the
aforementioned potential causes---security environment, economy, domestic
politics, and ideology / culture---that the first is most important: ``choices about
mechanization are strongly associated with a state's security environment.''

\subsection{The case for agent-based modeling}

There is reason to believe that the orthodox multivariate statistical modeling 
approach employed in \citet{sechser2010army} is insufficient and leads to
incorrect conclusions.

One, from a statistical perspective, the ability of the model in
\citet{sechser2010army} to produce reliable insight is questionable.
Observational data. Selection into war and deterrence.

@@

@@

\end{document}

>>>>

	Military mechanization is the degree to which a state's army is comprised of
	ground combat vehicles as opposed to infantry. Mechanization is both a 
	theoretically important topic in international relations scholarship and a practical 
	concern for policymakers. \citet{sechser2010army} influentially argue that
	inter-state dynamics (e.g., conflict) better explain mechanization choices than do
	intra-state factors (e.g., culture). This paper argues for the
	necessity of agent-based modeling of mechanization. ``Agentizing'' the
	model of \citet{sechser2010army} shows that their findings do not hold.
	An alternate model that treats mechanization as a complex system is presented 
	and discussed.

@@

@@

This paper presents an agent-based model of military mechanization. It
reviews the academic literature on military mechanization. It discusses why
orthodox modeling strategies fail to fully describe international security competition
as a complex system and why agent-based modeling is needed. In support of these
claims, it ``agentizes'' the leading model of military mechanization and shows
that it does not hold. Finally, it presents a
novel agent-based model that better represents military mechanization.

@@

There are many potential causes of mechanization. Inter-state
competition and war is a prominent candidate. States may choose choose mechanization policies 
that they believe will allow them to prevail against potential adversaries, or to send a
signal of deterrence, or to learn from the perceived mistakes of the last war
\citep{mearsheimer1983conventional,huth1988extended,murray2011military}. 
Economic endowments may play a role: e.g., capital-rich states may mechanize more
than labor-rich states \citep{gartzke2001democracy}. Domestic political
institutions may matter: e.g., autocracies may be more willing than democracies to adopt
low-mechanization ``human wave'' tactics
\citep{quinlivan1999coup,reiter2002democracies,brooks2008shaping,talmadge2015dictator}. 
Ideology \citep{van1984cult} and national culture \citep[e.g.][]{pollack2004arabs} are
hard-to-measure but perhaps influential. Lastly,
organizational culture can play a role \citep{builder1989masks}. Militaries may pursue a
certain mechanization posture because it aligns with their institutional
objectives, which may or may not be coincident with rational strategy. 
Consider Ireland: between the 1920s and 1940s, it seriously considered expending a 
large portion of its defense budget on building a relatively small number of tanks. 
They would have been wholly insufficient to repel a British invasion (their ostensible purpose), 
and Ireland would have been better-served to pursue a low-mechanization guerrilla defense strategy, 
but tanks were seen as a prestigious signifier of a ``real'' professional
military and hence desirable
\citep{farrell1998professionalization,farrell2001transnational}.

\citet{sechser2010army} is the landmark study on military mechanization. They
assemble a longitudinal dataset at the country-year level of the mechanization
of all states between 1979 and 2001. They conduct regression analysis on this
panel data. They use a ordinary least squares (OLS) linear regression model:

\begin{equation}
y_{ij} = \beta_0 + \sum_{k=1}^P \beta_{ijk}x_{ijk} + \epsilon
\end{equation}

with $y_{ij}$ the mechanization level of state $i$ in year $j$ (measured as the 
ratio of ground combat vehicles to infantry in a state's army) and $x_1 \ldots
x_j \ldots x_{P}$ as covariates hypothesized to be associated with
mechanization. They conclude that out of the
aforementioned potential causes---security environment, economy, domestic
politics, and ideology / culture---that the first is most important: ``choices about
mechanization are strongly associated with a state's security environment.''

\section{The case for agent-based modeling}

Although \citet{sechser2010army} is a valuable contribution to the literature,
its orthodox multivariate statistical modeling is insufficient. Agent-based
modeling is needed to capture mechanization as a complex system. These two
claims are addressed in turn.

\subsection{Inadequacy of statistical modeling}

\citet{sechser2010army} operate on observational rather than experimental
data. The assignment of treatments to subjects is not randomized, causing
\textit{selection bias}. For example, \citet{sechser2010army} operationalize
dyadic ``rivalry'' as whether two states recently engaged in a Militarized
Interstate Dispute (MID) \citep{cow_midb}, and find significant association with mechanization
(states match their mechanization level to that of their enemy). But interstate
conflict is not randomly assigned; rather, at least one state must select
themselves into it \citep{fearon2002selection}. States differ in their
proclivity for conflict initiation based on many factors, e.g. certain kinds of autocracies may be
especially war-prone \citep{de1995war}. Thus, correlation attributed to rivalry is in fact
bundling in many uncontrolled-for covariates, casting doubt on the claimed
finding. Natural experiments---a situation in which assignation of treatment is
known to be random---are the typical econometric strategy to avoid bias. For
example, \citet{angrist2002vouchers} exploit the fact that private school
vouchers were distributed by lottery in Colombia to obtain unbiased estimates of
the effect of schooling type (private vs. public) on educational achievement.
But such situations are rare in international relations---e.g., one cannot
even conceive of a way in which GDP per capita could be randomly assigned---and so
\cite{sechser2010army} lack such a control. Given rife selection effects and few
natural experiments, econometric modeling simply may not be an appropriate tool for mechanization.

The approach of \citet{sechser2010army} does not adequately represent complex interactions. First consider 
\textit{system decomposibility}. The additive linear model in (1) implicitly assumes 
that each covariate $x$ is separable and interpretation of the corresponding
coefficient $\beta$ depends on independence (i.e. a lack of multicollinearity).
It also assumes that mechanization $y$ is a function of those 
$x$. But evidence contradicts both assumptions. Take as an example GDP per
capita, which \citet{sechser2010army} find to be negatively correlated with
mechanization at the $p < 0.05$ level. We can represent this finding as a causal
diagram in the style of \citet{pearl2009causality}.\footnote{To be clear,
\citet{sechser2010army} studiously avoid claiming causation in their text. But
their choice in what is modeled as $y$ and what is modeled $x$ says a lot about the perceived
direction of association.}

But, a vast body of literature suggests that wealth is associated with
democracy. Causality appears to flow in both directions, i.e. democracy causes wealth and wealth also
causes democracy \citep{helliwell1994empirical,barro1996democracy,robinson2006economic,doucouliagos2008democracy}. 
Democracy affects conflict propensity \citep{maoz1993normative,russett1994grasping}. 
Conflict also affects democracy, e.g. a foreign-imposed regime change a la 1956
Hungary or 2003 Iraq
\citep{werner1996absolute,lo2008ensuring,downes2013forced}, and mechanization, 
as claimed by \citet{sechser2010army}. Mechanization affects battlefield success in war
\citep{biddle2004military,lyall2009rage}. War affects economic growth: 
consider the contrasting examples of WWII-era United States and present-day
Syria. If we represent these claims---which, it must be emphasized, do not even
scratch the surface of the potential system complexity---as a causal diagram, we
have:

Wealth is 
Interactions --> causal arrows, GDP per capita, democracy, war, mechanization
--> decomposibility of system

Path dependence

Cognition

ositive feedback loop

Path dependence

certain kinds of interactions

\subsection{Unreliable data}

Militarized Interstate Disputes Database (MIDB)
\citep{cow_midb}
to produce reliable insight is questionable.
Observational data. Selection into war and deterrence.

MID data sucks

@@

Lag effects = path dependence

Seeing what others are doing

@@

In summary, the orthodox view of mechanization is that it is a
reflexive process: states choose their mechanization posture largely based on
that neighbors, enemies, and recent military experiences (e.g. defeat in
counter-insurgency). This paper argues
that this view---based on traditional longitudinal data analysis---is ripe for
agentization. It implicitly argues for autonomous agents (states) that have
cognition regarding mechanization that takes into account other agents (e.g.
enemies), spatiality (e.g. neighbors), and learning from experience. This paper
builds such a model and validates it against the empirical record, using the
National Mechanization Index of \citet{sechser2010army}. It finds that the 
orthodox view of mechanization does not stand up to agentization. A positive feedback loop 
causes states to
mechanize much faster and to a greater degree than occurs in the real-world.
This implies the existence of some countervailing factor (possibly a negative
feedback loop) that mitigates the extent of mechanization. Candidates for what
this could be mostly center on ``internal'' factors---e.g., culture, economy,
and so on---as opposed to the ``external'', security environment-centric
orthodox view.
\end{document}
